---
title: "Introduction to ggseqplot"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to ggseqplot}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
csl: apa.csl
bibliography: references.bib
editor_options: 
  markdown: 
    wrap: 72
---


```{r prelude, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=8,
  fig.height=4.94
)

old <- options()        
on.exit(options(old))
options(rmarkdown.html_vignette.check_title = FALSE)

pkgs <- c("colorspace", "ggplot2", "ggthemes", "ggseqplot",
          "patchwork", "forcats", "ggh4x", "purrr", "TraMineR")

# Load all packages to library and adjust options
lapply(pkgs, library, character.only = TRUE)


```

# Prelude & Setup

Following Fasang and Liao [-@fasang2014], we distinguish between
**sequence representation** and **summarization** graphs. Summarization plots 
aggregate information from the data without displaying individual sequences. 
Given the complexity of sequence data, these plots focus on one or two 
dimensions of information stored in sequence data [@brzinsky-fay2014]. Common 
examples include transition plots, Kaplan-Meier curves, modal state plots, mean 
time plots, state distribution plots, and entropy plots [@fasang2014; @raab2022].

[`{ggseqplot}`](https://maraab23.github.io/ggseqplot/) includes five types of summarization plots:

-   State distribution plots (`ggseqdplot`)
-   Entropy line plots (`ggseqplot`)
-   Modal state sequence plot (`ggseqmsplot`)
-   Mean time plot (`ggseqmtplot`)
-   Transition rate plots (`ggseqtrplot`)

In contrast to summarization plots, representation plots visualize actually 
observed sequences. The most basic version is the sequence index plot, which 
displays all sequences in the dataset. However, this approach becomes impractical 
with large datasets due to overplotting—individual sequences may overlap, making 
the visualization misleading. To address this, alternative representations that 
display only a subset of sequences have been proposed.  

[`{ggseqplot}`](https://maraab23.github.io/ggseqplot/) supports both full and subset-based representation plots, offering the following plot types:

-   Sequence index plot (`ggseqiplot`)
-   Sequence frequency plot (`ggseqfplot`)
-   Representative sequence plot (`ggseqrplot`)
-   Relative frequency sequence plot (`ggseqrfplot`)

For a deeper discussion of sequence data visualization, see Brzinsky-Fay 
[-@brzinsky-fay2014], Fasang and Liao [-@fasang2014] and Chapter 2 of Raab & 
Struffolino [-@raab2022].

With the exception of transition rate plots, all of the plots listed
above can also be produced with
[`{TraMineR}`](http://traminer.unige.ch){target="_blank"}. In fact, 
[`{TraMineR}`](http://traminer.unige.ch){target="_blank"} typically offers more 
customization options—but only through base R's plotting system. If you are 
comfortable using base R graphics, there may be little reason to switch to [`{ggseqplot}`](https://maraab23.github.io/ggseqplot/).



However, [`{ggseqplot}`](https://maraab23.github.io/ggseqplot/) was developed 
precisely because many R users today prefer 
[`{ggplot2}`](https://ggplot2.tidyverse.org/){target="_blank"} to base
R's `plot` environment for visualizing data.
[`{TraMineR}`](http://traminer.unige.ch){target="_blank"}
[@gabadinho2011] was developed before
[`{ggplot2}`](https://ggplot2.tidyverse.org/){target="_blank"}
[@wickham2016] was as popular as it is today and back then many users
were more familiar with coding base R plots. Today, most researchers and students are more familiar with [`{ggplot2}`](https://ggplot2.tidyverse.org/){target="_blank"} and would rather 
build on existing skills instead of learning how to refine base R plots just for sequence visualizations.

This vignette demonstrates how sequence objects created via
`TraMineR::seqdef` are reshaped to plot them as ggplot2-typed figures
using [`{ggseqplot}`](https://maraab23.github.io/ggseqplot/). 

It outlines which functions from [`{TraMineR}`](http://traminer.unige.ch){target="_blank"} and the [`{ggplot2}`](https://ggplot2.tidyverse.org/){target="_blank"} universe are involved in generating the visualizations.

The vignette further illustrates how the appearance of plots produced with [`{ggseqplot}`](https://maraab23.github.io/ggseqplot/) can be customized using
[`{ggplot2}`](https://ggplot2.tidyverse.org/){target="_blank"} functions
and extensions.


### Setup example

We begin by loading the required libraries and preparing the sequence data to be 
visualized. The examples are based on data sets and examples provided by the
[`{TraMineR}`](http://traminer.unige.ch){target="_blank"} package.

\

<details><summary>**Click to show code for installing and loading required packages**</summary>

```{r libraries, eval=FALSE}
## ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
## Load and download (if necessary) required packages ----
## ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~


## Save package names as a vector of strings
pkgs <- c("colorspace", # for using colors palettes
          "forcats", # for dropping unused factor levels with `fct_drop`
          "ggh4x", # for proportional panel sized with `force_panelsizes`
          "ggplot2", # for using all the ggplot2 functions 
          "ggthemes", # for getting access to the canva_palettes
          "patchwork", # for working with plot types built with patchwork
          "purrr", # used in the grouped rplot example
          "TraMineR") # the ultimate sequence analysis suite


## Install uninstalled packages
lapply(pkgs[!(pkgs %in% installed.packages())], 
       install.packages, repos = getOption("repos")["CRAN"])


## Load all packages to library and adjust options
lapply(pkgs, library, character.only = TRUE)

## Don't forget to load ggseqplot
library(ggseqplot)

```

</details>


```{r setup, message=FALSE}


## ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
## Creating state sequence objects from example data sets ----
## ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

## biofam data
data(biofam)

biofam.lab <- c("Parent", "Left", "Married", "Left+Marr",
                "Child", "Left+Child", "Left+Marr+Child", "Divorced")

biofam.seq <- seqdef(biofam[501:600, ], 10:25, # we only use a subsample
                     labels = biofam.lab,
                     weights = biofam$wp00tbgs[501:600])

## actcal data
data(actcal)

actcal.lab <- c("> 37 hours", "19-36 hours", "1-18 hours", "no work")

actcal.seq <- seqdef(actcal,13:24,
                     labels=actcal.lab)

## ex1 data
data(ex1)
ex1.seq <- seqdef(ex1, 1:13, 
                  weights=ex1$weights)

```

**Note:** The default figure size in this document is 
`fig.width=8, fig.height=4.94`
<br><br><br>

# Technicalities

All functions in [`{ggseqplot}`](https://maraab23.github.io/ggseqplot/) follow a 
similar structure: they take a state sequence object—created with 
`TraMineR::seqdef`—as their starting point. The functions either operate directly 
on the sequence data or use other [`{TraMineR}`](http://traminer.unige.ch){target="_blank"} functions, such as `TraMineR::seqstatd`, to derive the information to be visualized.

Internally, [`{ggseqplot}`](https://maraab23.github.io/ggseqplot/) reshapes the 
data into a tidy (long) format before plotting them with [`{ggplot2}`](https://ggplot2.tidyverse.org/){target="_blank"} functions. This 
reshaping step is typically necessary for creating layered and faceted graphics 
in the grammar of graphics framework.

The following example demonstrates this process using a state distribution plot. 
First, we compute the cross-sectional state distributions across the sequence 
positions:

```{r seqstatd}
seqstatd(actcal.seq)
```

When calling `ggseqdplot`, these distributional data are internally reshaped 
into a long-format dataset, where each row represents the (weighted) relative 
frequency of a specific state at a specific sequence position.

For example, the `actcal.seq` data contain sequences of length 12 with an 
alphabet of 4 states. The reshaped dataset used for plotting will therefore 
consist of $12 \times 4 = 48$ rows. If a grouping variable is provided, the 
resulting dataset will contain 48 rows per group. To inspect this internal data 
structure, assign the plot output to an object. The resulting list object stores 
the data as its first element (named `data`).

```{r dplot-data}
dplot <- ggseqdplot(actcal.seq)
dplot$data
```

Once the data are in the correct shape, [`{ggseqplot}`](https://maraab23.github.io/ggseqplot/) 
uses standard [`{ggplot2}`](https://ggplot2.tidyverse.org/){target="_blank"} 
functions to create the visualizations. In the case of state distribution plots, 
for instance, `ggseqdplot` renders stacked bar charts for each sequence position 
using `ggplot2::geom_bar`.

The following table summarizes the core internal function calls for the various plot types available in [`{ggseqplot}`](https://maraab23.github.io/ggseqplot/):

| ggseqplot function | TraMineR function                                  | ggplot2 geoms and extensions                                                                                            |
|:--------------|:---------------|:----------------------------------------|
| `ggseqdplot`         | `TraMineR::seqstatd`                                         | `ggplot2::geom_bar`<br>optional: `geom_line`                                                                                            |
| `ggseqeplot`          | `TraMineR::seqstatd`                                         | `ggplot2::geom_line`                                                                                                                    |
| `ggseqmsplot`        | `TraMineR::seqmodst`                                         | `ggplot2::geom_bar`                                                                                                                     |
| `ggseqmtplot`        | `TraMineR::seqmeant`                                         | `ggplot2::geom_bar`                                                                                                                     |
| `ggseqtrplot`        | `TraMineR::seqtrate`                                         | `ggplot2::geom_tile`                                                                                                                    |
| `ggseqiplot`         | `TraMineR::seqformat`                                        | `ggplot2::geom_rect`                                                                                                                    |
| `ggseqfplot`         | `TraMineR::seqtab`                                           | `ggplot2::geom_rect`<br>[`{ggh4x}`](https://teunbrand.github.io/ggh4x/){target="_blank"} (for the axis labeling if group has been specified)                                                   |
| `ggseqrplot`         | `TraMineR::seqrep`                                           | `ggplot2::geom_rect`<br>`ggrepel::geom_text_repel`<br>[`{ggtext}`](https://wilkelab.org/ggtext/){target="_blank"} (for optional colored axis labels)<br>[`{patchwork}`](https://patchwork.data-imaginist.com/index.html){target="_blank"} (to combine plots) |
| `ggseqrfplot`        | `TraMineR::seqrfplot` | `ggplot2::geom_rect`<br>`ggplot2::geom_boxplot`<br>[`{patchwork}`](https://patchwork.data-imaginist.com/index.html){target="_blank"} (to combine plots)                                                              |

Most plots created with [`{ggseqplot}`](https://maraab23.github.io/ggseqplot/) 
behave like regular ggplots. You can customize them by adding themes, adjusting 
scales, or using other standard [`{ggplot2}`](https://ggplot2.tidyverse.org/){target="_blank"} syntax via `+`.

Representative sequence plots and relative frequency sequence plots, however, 
behave differently because they are composed of multiple subplots which are 
arranged using the 
[`{patchwork}`](https://patchwork.data-imaginist.com/index.html){target="_blank"} 
library. The following section illustrates how the appearance of the plots can
be changed.
<br><br><br>  

# Changing the appearance of plots

## The default case
As mentioned above, most plots rendered by 
[`{ggseqplot}`](https://maraab23.github.io/ggseqplot/) are of class 
`c("gg", "ggplot")` and can be customized  just like any other plot created with 
[`{ggplot2}`](https://ggplot2.tidyverse.org/){target="_blank"} 


### Example 1: State distribution plot

We begin with a simple state distribution plot using the sequence object 
`actcal.seq`, without modifying the default settings. 

```{r, dplot1}
ggseqdplot(actcal.seq)
```

Next, we demonstrate how to refine the plot using a variety of
[`{ggplot2}`](https://ggplot2.tidyverse.org/){target="_blank"} functions
and extensions. As with any other 
[`{ggplot2}`](https://ggplot2.tidyverse.org/){target="_blank"}
plot, the appearance of plots generated with 
[`{ggseqplot}`](https://maraab23.github.io/ggseqplot/) functions
can be  extensively customized:

```{r, message=FALSE, warning=FALSE}
ggseqdplot(actcal.seq) +
  scale_fill_discrete_sequential("heat") +
  scale_x_discrete(labels = month.abb) +
  labs(title = "State distribution plot",
       x = "Month") +
  guides(fill=guide_legend(title="Alphabet")) +
  theme_minimal(base_family = "") +
  theme(plot.title = element_text(size = 30, margin=margin(0,0,20,0)),
        plot.title.position = "plot",
        axis.title.x = element_text(hjust = 1, 
                                    vjust = -1),
        axis.title.y = element_text(hjust = 1, 
                                    vjust = 1),
        axis.title = element_text(size = 9))
```

In the following example, we use two `ggseqdplot` calls to create plots of the 
same data, one with and one without weights. We also display entropy lines 
(`geom_line`) and combine the two plots using the
[`{patchwork}`](https://patchwork.data-imaginist.com/index.html){target="_blank"} 
package [@pedersen2020].

```{r, message=FALSE, warning=FALSE}
# Save plot using weights
p1 <- ggseqdplot(ex1.seq, 
                 with.entropy = TRUE) + 
  ggtitle("Weighted data")

# Save same plot without using weights
p2 <- ggseqdplot(ex1.seq, 
                 with.entropy = TRUE,
                 weighted = FALSE) + 
  ggtitle("Unweighted data")

# Arrange and refine plots using patchwork
p1 + p2 + 
  plot_layout(guides = "collect") &
  scale_fill_manual(values= canva_palettes$`Fun and tropical`[1:4]) &
  theme_minimal(base_family = "") &
  theme(plot.title = element_text(size = 20,
                                  hjust = 0.5),
        legend.position = "bottom",
        legend.title = element_blank(),
        axis.title.x = element_text(hjust = 1, 
                                    vjust = -1),
        axis.title.y = element_text(hjust = 1, 
                                    vjust = 1),
        axis.title = element_text(size = 9))
```


### Example 2: Transition rate plot

In the next set of examples, we refine plots generated by `ggseqtrplot`, which 
internally calls `TraMineR::seqtrate` to compute the transition rates between 
states. These rates are returned in a symmetrical matrix and then reshaped into 
long format within `ggseqtrplot` to be plotted with `ggplot2::geom_tile`.

We begin with a basic example that visualizes transition rate matrices for two 
groups defined by the sex variable The output is a faceted plot visualizing two
transition rate matrices of DSS sequence data.

```{r}
ggseqtrplot(actcal.seq, 
            group = actcal$sex)
```

In the second example, we add customizations and utilize once
again the [`{patchwork}`](https://patchwork.data-imaginist.com/index.html){target="_blank"} 
library to compose a figure that compares the transition matrices of the same 
sequence data in STS and DSS format.

We set `x_n.dodge = 2` to prevent overlapping x-axis labels, reduce the font 
size of the value labels displayed within the tiles (`labsize = 7`), 
and set `dss = FALSE` to switch to the STS format.

```{r}

p1 <- ggseqtrplot(biofam.seq, 
                  dss = FALSE, 
                  x_n.dodge = 2,
                  labsize = 7) +
  ggtitle("STS Sequences") +
  theme(plot.margin = unit(c(5,10,5,5), "points"))

p2 <- ggseqtrplot(biofam.seq, 
                  x_n.dodge = 2,
                  labsize = 7) +
    ggtitle("DSS Sequences") +
  theme(plot.margin = unit(c(5,5,5,10), "points"))

p1 + p2 &
  theme(plot.title = element_text(size = 20,
                                  hjust = 0.5))

```

In contrast to the grouped plot above, the combined figure contains duplicated 
y-axis labels and titles. These can be removed using standard `ggplot2::theme()` arguments:

```{r}

p2 <- p2 +
  theme(axis.text.y = element_blank(), 
        axis.title.y = element_blank())

p1 + p2 &
  theme(plot.title = element_text(size = 20,
                                  hjust = 0.5))
```

### Example 3: Sequence index plots of illustrative (very few) sequences 

[`{TraMineR}`](http://traminer.unige.ch){target="_blank"} offers two types of sequence index plots: `TraMineR::seqiplot` for displaying a small selection of sequences (typically no more than 10), and `TraMineR::seqIplot` for plotting the full set of sequences. The former adds white space between the displayed sequences, making it well suited for illustrative purposes—for example, when showcasing a few hand-picked sequences in publications or teaching materials. While [`{ggseqplot}`](https://maraab23.github.io/ggseqplot/) does not provide built-in spacing between sequences, a similar effect can be achieved with a small trick: by assigning each sequence to its own group via the group argument, `ggseqiplot` uses `ggplot2::facet_wrap` to place each sequence in a separate panel, thereby visually separating them.

In the example below, we display five sequences (`biofam.seq[2:6,]`) as a faceted plot, where each sequence forms its own group (`group = 1:5`) and is shown in a separate panel arranged in a single column (`facet_ncol = 1`). The remainder of the code applies visual fine-tuning: group labels are moved to the left (`strip.position = "left"`) and rotated (`strip.text.y.left = element_text(angle = 0)`); superfluous axis elements—such as y-axis labels (`axis.text.y = element_blank()`), ticks (`axis.ticks.y = element_blank()`), and the axis title (`labs(y = NULL)`)—are removed; and the white space between sequences is slightly reduced by adjusting vertical panel spacing (`panel.spacing.y = unit(1.5, "lines")`).


```{r}
# plot a subset of 5 sequences from biofam.seq
ggseqiplot(biofam.seq[2:6,],
           group = 1:5,
           facet_ncol = 1,
           strip.position = "left",
           no.n = TRUE,
           border = TRUE, 
           weighted = FALSE) +
  labs(y = NULL) +
  theme(strip.text.y.left = element_text(angle = 0),
        panel.spacing.y = unit(1.5, "lines"),
        axis.text.y =  element_blank(),
        axis.ticks.y = element_blank())
```


### Example 4: Flipping coordinates

Another way to adjust the appearance of plots rendered by
[`{ggseqplot}`](https://maraab23.github.io/ggseqplot/) is to flip their 
coordinates—an approach commonly used in the [`{ggplot2}`](https://ggplot2.tidyverse.org/){target="_blank"} 
ecosystem.

In the example below, we illustrate coordinate flipping for both a mean time 
plot and a sequence index plot. For each plot type, we show the default version 
followed by its flipped counterpart:

```{r}
## default plot
ggseqmtplot(actcal.seq, no.n = TRUE, error.bar = "SE") 

## flipped version
ggseqmtplot(actcal.seq, no.n = TRUE, error.bar = "SE") +
 coord_flip() +
 theme(axis.text.y=element_blank(),
       axis.ticks.y = element_blank(),
       panel.grid.major.y = element_blank(),
       legend.position = "top")
```

While the flipped version above may be more in line with some users’ aesthetic 
preferences, flipping coordinates for sequence index plots is a more opinionated 
design choice.  Most researchers display time along the horizontal axis. However, 
if you prefer time to *run*  vertically—bottom to top, as in Piccarreta and 
Lior [@piccarreta2010]—this preference is easily accommodated: 

```{r}
## default plot
ggseqiplot(actcal.seq, sortv = "from.end") + 
  scale_x_discrete(labels = month.abb) 

## flipped version
ggseqiplot(actcal.seq, sortv = "from.end") + 
  scale_x_discrete(labels = month.abb) +
  coord_flip()

```


### Example 5: Proportional panel sizes for grouped sequence plots

A common use case—especially in the context of cluster analyses—is to display grouped sequence plots with panel sizes that reflect the relative size of each group. This makes the visualization more informative, as it visually communicates the prevalence of each group.

In the following example, we illustrate this approach using `actcal$sex` as the grouping variable. As before, the `group` argument is used to split the data into separate panels arranged in a single column (`facet_ncol = 1`). To scale the height of each panel proportionally to the group size, we compute the relative frequencies using:

```{r}
hghts <- table(fct_drop(actcal$sex)) / nrow(actcal.seq)
```

The `fct_drop()` function ensures that only those factor levels actually present in the data are considered when tabulating group sizes—this avoids reserving space for empty panels corresponding to unused levels. We then pass the resulting proportions to `ggh4x::force_panelsizes()` via the `rows` argument. This function adjusts the panel heights accordingly, so that larger groups occupy more vertical space in the plot. Finally, we reduce the spacing between panels using `theme(panel.spacing = unit(1, "lines"))` for a more compact appearance.



```{r}
## use ggh4x::force_panelsizes to get proportional panels
ggseqiplot(actcal.seq, 
           sortv = "from.end", 
           group = actcal$sex,
           facet_ncol = 1) +
  force_panelsizes(rows = hghts) +
  theme(panel.spacing = unit(1, "lines"))

```



## The special case of combined plots

Two types of plots differ from the standard behavior of
[`{ggseqplot}`](https://maraab23.github.io/ggseqplot/) functions: they
are composed by two subplots, which are arranged into a single figure using the 
[`{patchwork}`](https://patchwork.data-imaginist.com/index.html){target="_blank"} 
library. Because of this composition, their appearance cannot be adjusted in exactly the same way as plots returned directly as `ggplot` objects. For more details on the functionality and syntax of
[`{patchwork}`](https://patchwork.data-imaginist.com/index.html){target="_blank"},  
we refer readers to the to the package documentation. 

### Example 6: Annotation and themes

Despite their special structure, many aspects of combined
[`{patchwork}`](https://patchwork.data-imaginist.com/index.html){target="_blank"}
plots can still be customized in ways familiar from
[`{ggplot2}`](https://ggplot2.tidyverse.org/){target="_blank"}. For example, 
themes and plot annotations can be modified using the `&` operator 
(instead of `+`, which is used for standard `ggplot` objects).


```{r}
## compute dissimilarity matrix required for plot
diss <- seqdist(biofam.seq, method = "LCS")

## Relative Frequency Sequence Plot
## default version
ggseqrfplot(biofam.seq, diss = diss, k = 12) 

## adjusted version
ggseqrfplot(biofam.seq, diss = diss, k = 12) &
  theme_minimal(base_family = "") &
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        plot.title = element_text(size = 12),
        axis.title.x = element_text(hjust = 1, 
                                    vjust = -1),
        axis.title.y = element_text(hjust = 1, 
                                    vjust = 1),
        axis.title = element_text(size = 9)) &
  plot_annotation(title = "Relative Frequency Sequence Plot")
```

If you want to manipulate the appearance of a specific subplot, however, the default
approach might not work as expected. For example, if you aim to change the x-axis labels
of the index plot, the following code will *not* yield the intended result:

```{r}
ggseqrfplot(biofam.seq, diss = diss, k = 12) +
  scale_x_discrete(labels = 15:30)
```

In this case, `scale_x_discrete()` will be applied to the boxplot (i.e., the last plot
constructed in the combined object), not to the index plot. This behavior arises because
the `+` operator affects the final `ggplot` object passed into the patchwork layout.

To adjust the appearance of individual subplots within a combined plot, you need to first
save the composite plot object, access its components, and then modify them
individually before reassembling the full plot.

```{r}
## save & view original plot
p <- ggseqrfplot(biofam.seq, diss = diss, k = 12)
p

## change appearance of sub-plots

## first component: index plot
p[[1]] <- p[[1]] +
  scale_x_discrete(labels = 15:30)

## second component: boxplot
p[[2]] <- p[[2]] + labs(title = "Changed title")

## adjusted plot
p
```

### Example 7: The complex case of grouped rplots

Things become a bit more complex when working with grouped representative sequence plots.
In such plots, each group contributes two subplots: one showing information on the
"quality" of the representative sequences (e.g., frequency, coverage, or centrality), and
another displaying the corresponding index plots.

If you want to adjust the x-axis labels, for example, you must specifically 
target the index plots, which appear in the second row of the combined layout. 
The subplots are arranged by row, so for a two-group example, the index plots 
correspond to subplots 3 and 4.

To apply the desired changes, you need to extract these components, modify them as needed,
and then re-plot the final figure.


```{r}
## Compute a pairwise dissimilarity matrix
diss <- seqdist(actcal.seq, method = "LCS")

## original plot
p <- ggseqrplot(actcal.seq, 
                diss = diss,
                nrep = 3,
                group = actcal$sex)
p

## adjusted sequence index subplots
p[[3]] <- p[[3]] +
  scale_x_discrete(labels = month.abb) 

p[[4]] <- p[[4]] +
  scale_x_discrete(labels = month.abb) 

p

```

### Example 8: What about grouped rfplots?

Grouped representative frequency plots are currently **not implemented** in 
`ggseqrfplot` and must be assembled *manually* using the [`{patchwork}`](https://patchwork.data-imaginist.com/index.html){target="_blank"} 
library.

In the following example, we: 

1. Create rfplots for each group (here: "Men" and "Women"),  
2. Slightly *misuse* [`{patchwork}`](https://patchwork.data-imaginist.com/index.html){target="_blank"}'s tag annotation to assign group-specific labels to the first plot in each row,
3. Remove the legend from the upper plot panel (`p$man`), and
4. Combine the two plots in a final stacked figure using a nested patchwork layout. 

Technically,  the resulting plot is a nested patchwork plot. According to the  [`{patchwork}`](https://patchwork.data-imaginist.com/index.html){target="_blank"} 
documentation,

>[i]t is important to note that plot annotations only have an effect on the top-level patchwork. Any annotation added to nested patchworks are (currently) lost. If you need to have annotations for a nested patchwork you’ll need to wrap it in wrap_elements() with the side-effect that alignment no longer works.^[https://patchwork.data-imaginist.com/articles/guides/annotation.html#titles-subtitles-and-captions]

For this reason, we refrained from using `patchwork::plot_annotation` or `ggplot2::ggtitle` to add group titles and used plot tags instead.

```{r}

diss <- seqdist(biofam.seq, method = "LCS")

sex <- biofam[501:600, "sex"]

p <- map2(
  levels(sex), # input x
  c("Men", "Women"), # input y
  function(x, y) {
    p <- ggseqrfplot(biofam.seq[sex == x,],
                     diss = diss[sex == x,sex == x],
                     k = 12)
    p[[1]] <- p[[1]] + labs(tag = y)
    return(p)
    }
  )

names(p) <- levels(sex)

(p$man & theme(legend.position = "none")) / p$woman

    
```

This approach gives you full control over the layout and design of grouped rfplots, 
even in the absence of direct support in the `ggseqrfplot` function.



------------------------------------------------------------------------

# References
